---
title: "Using CSMatch in Practice: The LaLonde Empirical Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using CSMatch in Practice: The LaLonde Empirical Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
options(dplyr.summarise.inform = FALSE)
```

This vignette walks through fitting Caliper Synthetic Matching (CSM) to
a real, and unusually difficult, dataset: the LaLonde job-training data.
The goal here is not to re-derive the method (see the paper, Che, Meng,
& Miratrix, 2026, *Caliper Synthetic Matching*, for that) but to show
the sequence of decisions an analyst actually makes when using the
`CSMatch` package -- choosing a distance metric and caliper, checking
match quality, comparing weighting schemes, and finally estimating an
effect -- and which package functions support each decision.

This vignette is a version of the supplement of the main paper (with some text quoted verbatum), and the code is adapted from the scripts in `scripts/lalonde-analysis/` in the package's GitHub repository.
See that directory for data cleaning code as well.

```{r setup}
library(CSMatch)
library(dplyr)
library(ggplot2)
```

**AI Declaration:** This summary vignette was generated by Claude AI from the aforementioned materials, and then reviewed and edited by Luke Miratrix.


## The data

The LaLonde data are a classic *within-study comparison*: we take the
**control group** of the National Supported Work Demonstration (NSWD),
a randomized job-training experiment, and match it to a large pool of
possible comparison units drawn from the Current Population Survey
(CPS). Because both groups are, in truth, "controls" (neither received
the job-training program), a good matching method should estimate an
effect close to 0. Any nonzero estimate is evidence of residual bias
from imperfect matching.

This is a genuinely hard matching problem: NSWD participants look
very different from the general CPS population (much lower prior
earnings, different demographics), so good matches are scarce for many
units. We use this dataset to illustrate not just the mechanics of the
package but also the trade-offs analysts have to navigate when overlap
is poor.
(We ignore issues of whether there is any reasonable comparability at all, e.g., consistency in measurement of variables, and we are ignoring also whether selection on observables is likely to hold here. That is what the WSC will ascertain.)

```{r}
lalonde_df <- readRDS(
  system.file("extdata", "lalonde_for_analysis.rds", package = "CSMatch")
)
dim(lalonde_df)
table(lalonde_df$Z)
```

The treatment indicator is `Z` (1 = NSWD, 0 = CPS) and the outcome `Y`
is 1978 earnings. Covariates `X1`-`X8` are indicators for race, ethnicity, marital status, and lack of a degree,
plus age, years of education, and earnings in 1974 and 1975.

## Deciding which covariates matter

Before touching the outcome, we explore which covariates are most
related to treatment assignment, so we know where to focus our caliper
choices. A simple propensity score model is enough for this:

```{r}
lalonde_df <- lalonde_df %>%
  mutate(
    X1 = as.numeric(X1), X2 = as.numeric(X2),
    X3 = as.numeric(X3), X4 = as.numeric(X4),
    COLL = ifelse(X6 >= 16, 1, 0)  # college completion indicator
  )

ps_fit <- glm(Z ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8,
              data = lalonde_df, family = binomial())

lalonde_df$logit_p <- predict(ps_fit, type = "link")

ggplot( lalonde_df, aes( logit_p, group=Z, col=as.factor(Z) ) ) +
  geom_density() +
  theme_minimal()
```

Looking at standardized coefficients (see paper supplement; not shown here) flags race,
Hispanic ethnicity, having a college degree, and 1975 earnings as the
most predictive covariates.
Also, overlap on the propensity score is
poor: the vast majority of CPS control units look nothing like a
typical NSWD participant.

This tells us two things: (1) we should make
sure our distance metric weights the identified variables appropriately, and
(2) we should expect many treated units to have no good match at all,
which the caliper mechanics below are designed to handle gracefully.

We carry two derived covariates forward into the matching step
alongside the original `X1`-`X8`: `COLL` (college completion) and
`logit_p` (the propensity score on the logit scale), which acts as a
one-dimensional summary of everything else in the propensity model.

## Choosing a distance metric and caliper

`get_cal_matches()` needs a **distance metric**, a per-covariate
**scaling**, and an initial **caliper**. These jointly define what
"close" means. We use the scaled $L_\infty$ ("maximum") distance,
which behaves like Coarsened Exact Matching: two units are within
caliper $c$ only if *every* scaled covariate difference is within $c$.

The scaling vector is where the substantive decisions live. We:

- Exactly match on the binary covariates (race, ethnicity, marital
  status, degree status) by giving them a very large scaling constant,
  so any mismatch on them blows past the caliper.
- Set a caliper of 3 years on age and 2 years on education by scaling
  those covariates by $1/3$ and $1/2$ respectively -- i.e. a
  caliper of 1 on the scaled distance corresponds to a difference of at
  most 3 years of age (or 2 years of education).
- Set a caliper of \$5,000 on 1974 earnings and \$1,400 on 1975
  earnings (\$1,400 because 1975 earnings turned out to matter more,
  per the propensity model, so we ask for a tighter match on it).
- Add calipers of 0.5 on college completion and about 0.35 (0.2
  treatment-group standard deviations of the logit propensity score)
  on `logit_p`, so overall propensity similarity also constrains
  matches, without dominating the more interpretable, substantive
  covariates above.

```{r}
DIST_SCALING <- tibble(
  X1 = 1000,     # % Black (exact match)
  X2 = 1000,     # % Hispanic (exact match)
  X3 = 1000,     # % Married (exact match)
  X4 = 1000,     # % No degree (exact match)
  X5 = 1 / 3,    # Age (caliper of 3 years)
  X6 = 1 / 2,    # Years of education (caliper of 2 years)
  X7 = 1 / 5000, # 1974 earnings (caliper of $5,000)
  X8 = 1 / 1400, # 1975 earnings (caliper of $1,400)
  COLL = 1 / 0.5,     # College completion
  logit_p = 1 / 0.35  # Propensity score (logit scale)
)
```

A caliper of 1 on this scaled distance is what we will start with: under the infinity-norm distance, a
unit is a candidate match only if it is within _all_ of these
substantive tolerances at once.

## Fitting an initial match

`get_cal_matches()` does the matching and, with `est_method = "csm"`,
also solves for synthetic-control weights within each treated unit's
matched set. Because overlap is poor here, we use `rad_method =
"adaptive"`: treated units whose matched set is empty at the initial
caliper get their caliper *grown* just enough to capture at least one
control, rather than being dropped.

```{r}
lalonde_scm <- lalonde_df %>%
  get_cal_matches(
    form = Z ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + COLL + logit_p,
    caliper = 1,
    metric = "maximum",
    rad_method = "adaptive",
    scaling = DIST_SCALING,
    est_method = "csm"
  )

lalonde_scm
```

The printout tells us that around 10% of the 
treated units needed an adaptive caliper above 1 to find even one
control.

We call those treated units within the caliper "feasible" (they have a match within the
target caliper) and the remainder as "infeasible" (their
individual adaptive caliper had to grow above the target to get any controls). We will come back to
this feasible/infeasible split when estimating effects.

## Picking a caliper from the data

An initial caliper of 1 was really just a starting point (it is large
enough that everyone gets matched, one way or another). To pick a more
meaningful caliper, `caliper_distance_plot()` shows the distribution,
across treated units, of the distance to their 1st, 3rd, and 6th
closest control, and can back out the caliper needed to hit a target
coverage rate:

```{r, fig.height = 5, warning=FALSE}
cdp <- caliper_distance_plot(lalonde_scm, tops = c(1, 3, 6),
                              target_percentile = 2 / 3)
cdp
```

We ask for the caliper that would give about two-thirds of treated
units at least one match within caliper (a deliberately tight target,
chosen here so we get a good estimate on good fraction of the treated units. We can then do a
caliper sensitivity analysis to see how much it matters. See below). The plot both shows the needed value on the plot and also
returns the value as an attribute, so we can use it directly:

```{r}
new_caliper <- attr(cdp, "caliper")
new_caliper
```

We refit with `update_matches()`, which reuses every other setting
from the original call:

```{r}
lalonde_scm <- update_matches(lalonde_scm, lalonde_df, caliper = new_caliper)
lalonde_scm
```

We now have more infeasible matches because we are more aggressive with defining good match quality.


## Checking covariate balance

Before looking at any outcomes, we check whether the matched sample
achieves reasonable covariate balance -- this is a diagnostic step we
can iterate on freely, since it does not touch `Y` and so cannot
introduce outcome-driven bias ("fishing").

```{r}
covs_names <- c("Age", "Education", "Earnings (1974)", "Earnings (1975)")
love_plot(lalonde_scm, covs = c("X5", "X6", "X7", "X8"),
          covs_names = covs_names) +
  labs(title = "Covariate Balance (LaLonde)")
```

The love plot adds treated units beyond the feasible set one unit at a time, in order of increasing
adaptive caliper (i.e. easiest-to-match units first), and tracks the
running treated-vs-control mean difference on each covariate. A
roughly flat, near-zero line means balance does not deteriorate as we
absorb harder-to-match units.
Interestingly, balance _improves_ for this dataset as we include the hard to match units.
This is not what we would typically find, nor expect.

## Comparing weighting schemes before looking at outcomes

CSM's synthetic-control weighting is one of several ways to combine
the units inside a matched set; simple averaging (like CEM) and
1-nearest-neighbor are common alternatives. `sensitivity_table()` fits
all of them and reports effective sample size and match-quality
metrics side by side, which lets us reason about the bias/variance
trade-off *before* committing to a weighting scheme:

```{r}
sensitivity_table(lalonde_scm, include_distances = TRUE) %>%
  select(Estimate, N_T, N_C, ESS_C, mean_dist, median_dist, SE_star, SE_ratio, bias_ratio) %>%
  knitr::kable( digits = 2 )
```

`ESS_C` is the control group's effective sample size -- how many hypothetical
"independent" control units would give the same precision as the weighting scheme,
after accounting for reuse and unequal weights. `mean_dist`/`median_dist`
summarize how close each treated unit's (weighted) synthetic control
is to the treated unit itself, which relates to potential extrapolation
bias: the closer to 0, the less we are relying on the model
implicitly extrapolating between units. `SE_star` is a rule-of-thumb
pseudo-SE, $\sqrt{1/N_T + 1/\text{ESS}_C}$, useful for *relative*
comparisons across rows here, not as a standalone SE. `SE_ratio` and
`bias_ratio` express each method relative to the unweighted, raw
average (`*_raw`) as a baseline.

In this dataset, the synthetic weighting (`ATT`/`FATT`) meaningfully
reduces the average distance to the matched controls relative to raw
averaging, at the cost of a smaller effective sample size (more reuse
of a few good controls) and therefore a larger SE. Whether that
trade is worth it is a judgment call that depends on the application;
`sensitivity_table()` is meant to make that trade-off visible so the
decision is deliberate rather than a byproduct of a default setting.

## Estimating the effect

With the matching and weighting settled, `estimate_ATT()` computes the
point estimate and its standard error. We report both:

- The **SATT** (sample average treatment effect on the treated), using
  every treated unit, including the "infeasible" ones matched only via
  an enlarged adaptive caliper.
- The **FSATT** ("feasible" SATT), restricting to treated units that
  had a match within our chosen target caliper.

```{r}
estimate_ATT(lalonde_scm, outcome = "Y")
estimate_ATT(lalonde_scm, outcome = "Y", feasible_only = TRUE)
```

Recall the target here is 0 (we are matching NSWD controls to CPS
"controls"). Both estimates are statistically indistinguishable from
0 given their standard errors, which is reassuring given the within study comparison -- but the SATT and FSATT need not always
agree. When they diverge, it is worth asking why:

- If the infeasible treated units look systematically different from
  the feasible ones, the discrepancy may reflect real effect
  heterogeneity, in which case the SATT (using everyone) is the more
  generalizable estimate.
- If instead the SATT reflects bias from a steep drop-off in match quality for the
  infeasible units, the FSATT (using only well-matched units) may be
  the more trustworthy estimate, at the cost of generalizing to a
  narrower population.

We note that the standard errors are estimated with our recommended methods. Other options are possible.

## Checking robustness to the caliper choice

Because the caliper of `r round(new_caliper, 3)` was itself a choice,
it is worth checking how sensitive the ATT/FATT estimates are to it.
`caliper_sensitivity_table()` refits the match across a grid of
calipers (this is the most expensive step in this vignette -- we use a
coarser grid here, `R = 10`, than the `R = 30` used for the smoother curves
in the paper):

```{r}
cal_sens_tbl <- caliper_sensitivity_table(
  lalonde_scm, lalonde_df,
  R = 10, outcome = "Y",
  min_cal = 0, max_cal = 10.01
)

caliper_sensitivity_plot(cal_sens_tbl, focus = c("ATT", "FATT")) +
  facet_wrap(~feasible) +
  geom_vline(xintercept = max(lalonde_scm$adacalipers), lty = 3)
```

The dotted line marks the caliper large enough to make every treated
unit feasible (at which point the FSATT and SATT panels necessarily
agree). Watching how the CSM, 1-nearest-neighbor, and raw-average
lines move (or don't) as the caliper widens is a useful diagnostic:
here, the raw average estimate drifts steadily away from 0 as poorer
matches are absorbed, while CSM's synthetic weighting stays closer to
0 across the range -- evidence that the synthetic step is doing real
work to control extrapolation bias.

## Summary: the decisions, in order

1. **Explore covariate importance and overlap** (e.g. via a propensity
   model) before touching the matching call, to help determine where calipers
   need to bind tightly.
2. **Choose a distance metric and per-covariate scaling** that encodes
   substantive tolerances (e.g. "a 3-year age gap is as bad as a
   \$5,000 earnings gap"), using `get_cal_matches()`'s `scaling`
   argument. You can exact-match categorical values with a large scaling constant.
3. **Pick a caliper using `caliper_distance_plot()`** against a target
   coverage rate, and refit cheaply with `update_matches()` as that
   choice changes. (This step is not necessary if the initial work of what determines a "good match" is settled from the prior steps.)
4. **Check balance with `love_plot()`** and compare weighting schemes
   with `sensitivity_table()` -- all before looking at the outcome, so
   these choices can't be accused of fishing. You can tweak your distance measure and/or calipers to try and get good balance, if desired.
5. **Estimate both the SATT and FSATT with `estimate_ATT()`**, and use
   `caliper_sensitivity_table()`/`caliper_sensitivity_plot()` to see
   whether the conclusion is robust to the caliper choice or is instead
   an artifact of it.
